---
title: "RNA-seq of BR16 experiments"
subtitle: "Compare results between BR16 experiments"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  se_file_1: ./output/rnaseq/br16_dtc/se_GRCh38_RFP_mVenus_gene_qc.rds
  se_file_2: ./output/rnaseq/br16_bulk/se_gene.rds
  dge_file_1: ./output/rnaseq/br16_dtc/dge_hclusters/dge-edgeR_QLF.rds
  dge_file_2: ./output/rnaseq/br16_bulk/dge/dge-edgeR_QLF.rds
  gse_file_1: ./output/rnaseq/br16_dtc/dge_hclusters/gse_omnibus-edgeR_QLF.rds
  gse_file_2: ./output/rnaseq/br16_bulk/dge/gse_omnibus-edgeR_QLF.rds
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

# dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)
library(ggrastr)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/eigencorval.R')
source('./code/R-functions/subchunkify.R')
source('./code/R-functions/pairwise_terms_jc.r')

clean_msigdb_names <- function(x) x %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Define additional functions
```{r add-functions}
clean_sample_alias <- function(x) 
  gsub('o[0-9]*_[0-9]*-|_gDNA.*', '', x) %>% gsub(".*-MS", "MS", .) %>% gsub("_S[0-9]+_R[0-9].*", "", .) %>% gsub('__STARpass1', '', .) %>% gsub(".k2report", "", .) %>% gsub(".*-|.disambiguated.*", "", .)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Load data
```{r load-data}
se_sc <- readRDS(params$se_file_1)
dge_sc <- readRDS(params$dge_file_1)
gse_sc <- readRDS(params$gse_file_1)

se_b <- readRDS(params$se_file_2)
dge_b <- readRDS(params$dge_file_2)
gse_b <- readRDS(params$gse_file_2)
```

Load list of dormancy genes
```{r load-dormancy-genes}
dormancy_genes <- read_csv(file = './data/resources/list_of_dormancy_genes_MS_20230823.csv')
dormancy_genes_ar <- read_csv(file = './data/resources/list_of_dormancy_genes_AR_20231201.tsv.txt') %>% pull('Gene')
```


Start indexes of results
```{r start-gene-comp-list}
gene_comp_list <- list()
```




## p27--over--WT at 96h
Compare results from :

  - Xenograft scRNA-seq **Dormant over Cycling** defined after clustering
  - Bulk dox treatment experiment: **p27 over WT** adjusted by time

  
```{r p27-wt-conf}
bulk_index <- "p27-wt-96"


# Join DGE tables
data_sc <- dge_sc[['Dormant--over--Cycling']]$results %>% 
  dplyr::select(feature, gene_name, logFC, PValue, FDR, description) 

data_b <- dge_b[[bulk_index]]$results %>% 
  dplyr::select(feature, gene_name, logFC, PValue, FDR, description)

study_sign_palette <- c(`scRNA and bulk` = '#C85200', scRNA = '#18A188', bulk = 'grey50', none = 'grey90')

dge_joined<- data_sc %>% 
  left_join(data_b, by = c('feature', 'gene_name', 'description'), suffix = c('_sc', '_bulk')) %>% 
  mutate(
    FDR_bulk = ifelse(is.na(FDR_bulk), 1, FDR_bulk),
    study_sign = ifelse(FDR_sc < 0.05, 'scRNA', 'none'),
    study_sign = ifelse(FDR_bulk < 0.05, 'bulk', study_sign),
    study_sign = ifelse(FDR_sc < 0.05 & FDR_bulk < 0.05, 'scRNA and bulk', study_sign),
    study_sign = factor(study_sign, levels = rev(names(study_sign_palette))),
    FDR_mean = (FDR_sc + FDR_bulk) / 2,
    logFC_mean = (logFC_sc + logFC_bulk) / 2
  )

# Join GSE tables
data_sc <- gse_sc[['Dormant--over--Cycling']]$GSEA[['msigdb.c2.cp']]@result
data_b <- gse_b[[bulk_index]]$GSEA[['msigdb.c2.cp']]@result
gse_joined <- data_sc %>% 
  left_join(data_b, by = c('ID', 'Description'), suffix = c('_sc', '_bulk')) %>% 
  mutate(
    FDR_bulk = ifelse(is.na(p.adjust_bulk), 1, p.adjust_bulk),
    study_sign = ifelse(p.adjust_sc < 0.05, 'scRNA', 'none'),
    study_sign = ifelse(p.adjust_bulk < 0.05, 'bulk', study_sign),
    study_sign = ifelse(p.adjust_sc < 0.05 & p.adjust_bulk < 0.05, 'scRNA and bulk', study_sign),
    study_sign = factor(study_sign, levels = rev(names(study_sign_palette))),
    p.adjust_mean = (p.adjust_sc + p.adjust_bulk) / 2,
    NES_mean = (NES_sc + NES_bulk) / 2
  )
gse_joined_long <- rbind(
  data_sc %>% mutate(study = 'scRNA-seq'),
  data_b %>% mutate(study = 'bulk')
) 
```

### Differential gene expression

#### Table

Significant genes (FDR<0.05) in either scRNA-seq or bulk experiment
```{r p27-wt-deg-table}
dge_joined %>% 
  filter(FDR_sc < 0.05 | FDR_bulk < 0.05) %>%
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'Combined DEG results',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            )) %>% 
  formatRound(c('logFC_sc', 'logFC_bulk', 'logFC_mean')) %>% 
  formatSignif(c('FDR_sc', 'FDR_bulk', 'FDR_mean')) %>% 
  formatStyle(
    'study_sign',
    target = 'row',
    backgroundColor = styleEqual(c('scRNA and bulk', 'scRNA', 'bulk', 'none'), 
                                 study_sign_palette
                                 ) ) 

```

#### Fold-changes scatterplot
```{r p27-wt-logFC-correlation-shared_genes-plot, fig.width = 3.2, fig.asp = 0.7}
cor_res <- cor.test(dge_joined$logFC_sc, dge_joined$logFC_bulk, method = 'spearman')
cor_res_txt <- paste0("r = ", round(cor_res$estimate, 2),", p = ", format.pval(cor_res$p.value, digits = 2))

top_genes <- dge_joined %>% filter(study_sign == 'scRNA and bulk') %>% arrange(desc(abs(logFC_mean))) %>% head(20)

dge_joined %>% 
  ggplot(aes(logFC_sc, logFC_bulk, color = study_sign, label = gene_name)) +
  geom_point_rast(alpha = 0.6, pch = 16, size = 1) +
  scale_color_manual(values = study_sign_palette) +
  geom_abline(lty = 'dashed', color = 'gray60') +
  geom_hline(yintercept = 0, lty = 'dashed', color = 'gray60') +
  geom_vline(xintercept = 0, lty = 'dashed', color = 'gray60') +
  geom_text_repel(color = 'black', data = top_genes, min.segment.length = 0, seed = 42, box.padding = 0.25, max.overlaps = Inf, show.legend = FALSE, size = geom_text_size*0.8, segment.size = linewidth_default) +
  labs(
    x = expression(paste("lo", g[2],"(fold-change) scRNA-seq")),
    y = expression(paste("lo", g[2],"(fold-change) bulk")),
    color = 'Classification based\non FDR<0.05',
    caption = paste('Top-20 shared genes labelled\n', 'Spearman', cor_res_txt)
  )

```



#### Fold-changes scatterplot:  scRNA-seq genes

here we show the correlation only genes in scRNA-seq that have a FDR < 0.05
```{r p27-wt-logFC-correlation-scrna_genes-plot, fig.width = 3.2, fig.asp = 0.7}
use_cd <- dge_joined %>% filter(FDR_sc < 0.05)
cor_res <- cor.test(use_cd$logFC_sc, use_cd$logFC_b, method = 'spearman')
cor_res_txt <- paste0("r = ", round(cor_res$estimate, 2),", p = ", format.pval(cor_res$p.value, digits = 2))

use_cd %>% 
  ggplot(aes(logFC_sc, logFC_bulk, color = study_sign, label = gene_name)) +
  geom_point_rast(alpha = 0.6, pch = 16, size = 1.5) +
  geom_smooth(aes(logFC_sc, logFC_bulk), color = 'blue', method = 'gam') +
  scale_color_manual(values = study_sign_palette) +
  geom_abline(lty = 'dashed', color = 'gray60') +
  geom_hline(yintercept = 0, lty = 'dashed', color = 'gray60') +
  geom_vline(xintercept = 0, lty = 'dashed', color = 'gray60') +
  labs(
    x = expression(paste("lo", g[2],"(fold-change) scRNA-seq")),
    y = expression(paste("lo", g[2],"(fold-change) bulk")),
    color = 'Classification based\non FDR<0.05',
    caption = paste('Spearman', cor_res_txt, '\nBlue line shows GAM fit and shaded area its confidence interval')
  ) +
  theme(
    plot.caption = element_text(size = 6)
  )
```


### Functional enrichment

#### Table
Significant genes (FDR<0.05) in either scRNA-seq or bulk experiment
```{r p27-wt-gse-table}
gse_joined %>% 
  filter(p.adjust_sc < 0.05 | p.adjust_bulk < 0.05) %>%
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'Combined GSEA results',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            )) %>% 
  formatRound(c('NES_sc', 'NES_bulk', 'NES_mean')) %>% 
  formatSignif(c('p.adjust_sc', 'p.adjust_bulk', 'p.adjust_mean')) %>% 
  formatStyle(
    'study_sign',
    target = 'row',
    backgroundColor = styleEqual(c('scRNA and bulk', 'scRNA', 'bulk', 'none'), 
                                 study_sign_palette
                                 ) ) 

```


#### NES scatterplot
```{r p27-wt-gse-nes-correlation-shared_genes-plot, fig.width = 3.2, fig.asp = 0.7}
cor_res <- cor.test(gse_joined$NES_sc, gse_joined$NES_bulk, method = 'spearman')
cor_res_txt <- paste0("r = ", round(cor_res$estimate, 2),", p = ", format.pval(cor_res$p.value, digits = 2))

top_gset <- gse_joined %>% filter(study_sign == 'scRNA and bulk') %>% arrange(desc(abs(NES_mean))) %>% head(20)

gse_joined %>% 
  filter(!is.na(study_sign)) %>% 
  mutate(
    Description = clean_msigdb_names(Description),
    study_sign = factor(study_sign, levels = names(study_sign_palette))
  ) %>% 
  ggplot(aes(NES_sc, NES_bulk, color = study_sign, label = Description)) +
  geom_point_rast(alpha = 0.6, pch = 16, size = 1) +
  scale_color_manual(values = study_sign_palette) +
  geom_abline(lty = 'dashed', color = 'gray60') +
  geom_hline(yintercept = 0, lty = 'dashed', color = 'gray60') +
  geom_vline(xintercept = 0, lty = 'dashed', color = 'gray60') +
  # geom_text_repel(color = 'black', data = top_gset, min.segment.length = 0, seed = 42, box.padding = 0.25, max.overlaps = Inf, show.legend = FALSE, size = geom_text_size*0.8, segment.size = linewidth_default) +
  labs(
    x = "NES scRNA-seq",
    y = "NES bulk",
    color = 'Classification based\non Adjusted P < 0.05',
    # caption = paste('Top-20 shared genes labelled\n', 'Spearman', cor_res_txt),
    caption = paste('Spearman', cor_res_txt)
  )
```


#### Dotplot
Dotplot showing the top-20 downregulated and top-20 upregulated gene-sets in scRNA-seq, and its results in the bulk experiment
```{r p27-wt-gse-dotplot-down, fig.width = 7.2, fig.asp = 0.7}
use_id_down <- gse_joined %>% 
  filter(NES_sc < 0) %>%
  filter(p.adjust_sc < 0.05) %>% 
  arrange(p.adjust_sc) %>% 
  pull(Description) %>% 
  head(20)

use_id_top <- gse_joined %>% 
  filter(NES_sc > 0) %>%
  filter(p.adjust_sc < 0.05) %>% 
  arrange(p.adjust_sc) %>% 
  pull(Description) %>% 
  head(20)

use_id <- c(as.character(use_id_top), as.character(use_id_down))

use_linewidth <- 1/2.141959
use_data <- gse_joined_long %>% 
  filter(Description %in% use_id) %>% 
  mutate(
    Description = clean_msigdb_names(Description),
    Description = factor(Description, levels = unique(clean_msigdb_names(use_id))),
    study = factor(study, levels = c('scRNA-seq', 'bulk'))
  ) %>% 
  group_by(Description, study) %>% 
  slice_min(p.adjust, n = 1, with_ties = FALSE) %>%
  ungroup()

use_data %>% 
  ggplot(aes(NES, Description, color = NES, size = -log10(p.adjust))) +
  geom_point() +
  scale_y_discrete(labels =  default_labeller(70)) +
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.1))) +
  scale_color_gradient2(trans = "reverse") +
  scale_size_continuous(range = c(1.3, 3)) +
  facet_wrap(~study) +
  geom_vline(xintercept = 0, lty = 'dashed', color = 'gray60') +
  labs(
    x = 'Normalized enrichment score',
    y = NULL,
    color = 'NES',
    size = expression(-log[10](Adjusted~p-value))
  ) +
  theme_facet

```
