---
title: "Differential gene expression of BR16 xenograft samples"
subtitle: "Human reads from order p28634_o32926"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  data_dir: ./output/rnaseq/br16_dtc
  output_dir: ./output/rnaseq/br16_dtc/dge
  se_file: se_GRCh38_RFP_mVenus_gene_qc.rds
  min_counts: 3
  min_present_prop: 0.70
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 300, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/eigencorval.R')
source('./code/R-functions/subchunkify.R')
source('./code/R-functions/pairwise_terms_jc.r')

clean_msigdb_names <- function(x) x %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Define additional functions
```{r add-functions}
clean_sample_alias <- function(x) 
  gsub('o[0-9]*_[0-9]*-|_gDNA.*', '', x) %>% gsub(".*-MS", "MS", .) %>% gsub("_S[0-9]+_R[0-9].*", "", .) %>% gsub('__STARpass1', '', .) %>% gsub(".k2report", "", .) %>% gsub(".*-|.disambiguated.*", "", .)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Configuration for functional enrichment
```{r conf-gse_omnbus}
msigdb_vs <- 'v2026.1.Hs'
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0('./data/resources/MSigDB/', msigdb_vs, '/c2.cp.', msigdb_vs, '.symbols.gmt'),
  msigdb.c2.cp.reactome = paste0('./data/resources/MSigDB/', msigdb_vs, '/c2.cp.reactome.', msigdb_vs, '.symbols.gmt')
)
```

Load list of dormancy genes
```{r load-dormancy-genes}
dormancy_genes <- read_csv(file = './data/resources/list_of_dormancy_genes_MS_20230823.csv')
dormancy_genes_ar <- read_csv(file = './data/resources/list_of_dormancy_genes_AR_20231201.tsv.txt') %>% pull('Gene')
```


Load data
```{r load-data}
se <- readRDS(file.path(params$data_dir, params$se_file))
```



## Configure analyses

### Data wrangling

Modify sample annotation
```{r dw-coldata}
colData(se) <- colData(se) %>% data.frame %>% 
  mutate(
    ## dormancy_brightness
    dormancy_brightness = gsub("\\|.*", "", fluorescence),
    dormancy_brightness = ifelse(
      dormancy_brightness == '3_DTCs_Bright_and_Dormant', 
      'Mixed', 
      dormancy_brightness),
    dormancy_brightness = ifelse(
      dormancy_brightness == 'Bright_Dormant', 
      'Mixed', 
      dormancy_brightness),
    dormancy_brightness = ifelse(
      dormancy_brightness == 'Dim-Bright', 
      'Dim', 
      dormancy_brightness),
    dormancy_brightness = ifelse(
      dormancy_brightness == 'Very_Dim', 
      'Dim', 
      dormancy_brightness),
    dormancy_brightness = ifelse(
      dormancy_brightness == 'No_info_about_dormancy_state', 
      'Unk', 
      dormancy_brightness),
    ## dormancy_status
    dormancy_status = ifelse(
      dormancy_status == 'mixed_dormant', 
      'mixed', 
      dormancy_status),
    ## dormancy_brightness
    dormancy_brightness = factor(dormancy_brightness,
                                 levels = c('Bright', 
                                            'Dim', 
                                            'Mixed', 
                                            'Cycling', 
                                            'Unk')),
    dormancy_brightness = droplevels(dormancy_brightness),
    ## check dormancy status
    dormancy_check = ifelse(dormancy_check == 'no',
                            'Regular',
                            dormancy_check),
    dormancy_check = ifelse(grepl('Picked_on_G_Floor_CC', dormancy_check),
                            'Different microscope',
                            dormancy_check),
    dormancy_check = factor(dormancy_check) %>% relevel(ref = 'Regular'),
    ## sample_type
    sample_type_o=sample_type,
    sample_type = gsub(";wbc|;dn", "", sample_type),
    sample_type = factor(sample_type) %>% relevel(ref = 'dtc_single'),
    #time_months
    time_months = factor(time_months, levels = as.numeric(time_months) %>% unique %>% sort %>% as.character)
  ) %>% 
  DataFrame
```

Filtering out low-abundance features. Keeping those features with at least 10 counts in at least 70% of samples of the smallest group.
```{r dw-subset-feature-abundance}
use_rows <- filterByExpr(se, group=paste(se$sample_type, se$dormancy_status), min.count = 1, min.prop = 0.3)
se <- se[use_rows,]

# Update PerCellQCMetrics
colData(se) <- colData(se) %>% data.frame %>% dplyr::select(-(sum:subsets_Ribo_percent)) %>% DataFrame
se <- addPerCellQCMetrics(
  se,
  subsets=list(
    Mito=which(rowData(se)$is.mito),
    Ribo=which(rowData(se)$is.ribo)
  ),
  percent.top = c(1,5,10,20, 50, 100),
  threshold = 3
  )

# # Table of removed features
# table(use_rows) %>% 
#   data.frame %>% 
#   mutate(use_rows = ifelse(use_rows == TRUE, 'keep', 'remove')) %>% 
#   column_to_rownames('use_rows') %>% 
#   kbl(caption = 'Filtered features') %>%
#   kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

TMM normalization
```{r dw-tmm}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)
```

Cell cycle annotation with cyclone
```{r dw-cc-cyclone}
cc.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", 
    package="scran"))
res <- cyclone(se, cc.pairs)
se$cc_phase <- res$phases
se$cc_g1_score <- res$normalized.scores$G1
se$cc_s_score <- res$normalized.scores$S
se$cc_g2m_score <- res$normalized.scores$G2M
```

### Define a list of comparisons
```{r conf-comparisons}
x <- colData(se) %>% data.frame
comp_list <- list(
  `bright-o-cycling` = list(
    Bright = x %>% 
      filter(dormancy_brightness == 'Bright') %>% 
      pull(sample_alias) %>% 
      sort,
    Cycling = x %>% 
      filter(dormancy_brightness == 'Cycling') %>% 
      pull(sample_alias) %>% 
      sort
  )
)

comp_list_description <- data.frame(
  comparison = names(comp_list),
  comparison_long = 'Vector bright over cycling'
)

```

The following comparisons were performed

- Vector bright over cycling : bright over cycling, dim excluded

#### Table
```{r table-comparisons}
i = names(comp_list)[1]
comp_df <- foreach(i = names(comp_list), .combine = rbind) %do% {
  icomp <- comp_list[[i]]
  rbind(
    data.frame(
      comparison = i,
      group = names(icomp)[1],
      sample = icomp[[1]]
    ),
    data.frame(
      comparison = i,
      group = names(icomp)[2],
      sample = icomp[[2]]
    )
  )
}

comp_df %>% 
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'DEG : List of comparisons',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            ))
```


## Data generation

### Differential expression analysis
Run differential expression analysis using EdgeR QLF
```{r edgeR-QLF-run, eval = FALSE}
ensembl <- biomaRt::useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
gene_desc <- biomaRt::getBM(attributes=c('external_gene_name','description'),
                              filters = 'external_gene_name',
                              values = rowData(se)$gene_name, mart =ensembl) %>%
    dplyr::rename('gene_name' = 'external_gene_name') %>%
    unique

use_comp <-  comp_list[[1]]
dge_list <- foreach(use_comp = comp_list) %do%{
  se_cols <- se$sample_alias %in% unlist(use_comp)
  use_se <- se[,se_cols]
  use_se$group <- ifelse(
    use_se$sample_alias %in% use_comp[[1]],
    names(use_comp)[1],
    names(use_comp)[2]
  )
  
  if(is.null(use_comp$batch)) {
    dge <- edgeR_dge(
      use_se,
      # Design configuration for differential expression
      group_var =  'group',
      group_sample = names(use_comp)[1],
      group_ref = names(use_comp)[2],
      design_formula = "~ group",
      coef = 'last',
      # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
      assay_to_DGEList = 'counts',
      assay_to_row_filter = "counts",
      # Feature filtering parameters
      use_filterByExpr = TRUE,
      min_counts = params$min_counts,
      min_present_prop = params$min_present_prop,
      # EdgeR workflow configuration
      run_calcNormFactors = 'TMM',
      estimateDisp_robust = TRUE,
      estimateDisp_trend.method = "locfit",
      glm_approach = "QLF",
      # Output configuration
      adjust_method = 'BH'
      )
    } else if(use_comp$batch == 'donor') {
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        batch_vars = 'donor',
        design_formula = "~ donor + group",
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
        )
    }

  use_res <- dge$results %>%  left_join(., gene_desc)
  dge$results <- use_res %>% 
    filter(!duplicated(feature)) %>% 
    mutate(rownames = feature) %>% 
    column_to_rownames('rownames')
  
   return(dge)
}

names(dge_list) <- names(comp_list)
detach("package:biomaRt", unload=TRUE)
# Save objects
saveRDS(dge_list, file = file.path(params$output_dir, 'dge-edgeR_QLF.rds'))

###########
#   use_samples = NULL
#   # Desing configuration for differential expression
#   group_var = NULL
#   group_sample = NULL
#   group_ref = NULL
#   numeric_covar = NULL
#   batch_vars = NULL
#   design_formula = NULL
#   contrast = NULL
#   coef = NULL
#   # Conversion from SingleCellExpeiment to DGEList
#   spike_normalization = FALSE
#   assay_to_DGEList = 'counts'
#   assay_to_row_filter = "tpm"
#   use_colData = NULL
#   use_rowData = NULL
#   # Feature filtering parameters
#   use_filterByExpr = TRUE
#   min_counts = 1
#   min_present_prop = 0.40
#   # EdgeR workflow configuration
#   run_calcNormFactors = 'TMM'
#   estimateDisp_robust = FALSE
#   estimateDisp_trend.method = "locfit"
#   glmQLFit_robust = FALSE
#   glm_approach = "QLF"
#   # Output configuration
#   adjust_method = 'BH'
#   assays_from_SingleCellExperiment = NULL
# 
# object = se
#     # Desing configuration for differential expression
#     group_var =  condition_var
#     group_sample = use_comp[1]
#     group_ref = use_comp[2]
#     batch_vars = 'sample_type'
#     design_formula = paste("~ 0 +", condition_var," + sample_type")
#     coef = NULL
#     # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
#     assay_to_DGEList = 'counts'
#     assay_to_row_filter = "counts"
#     # Feature filtering parameters
#     use_filterByExpr = TRUE
#     min_counts = params$min_counts
#     min_present_prop = params$min_present_prop
#     # EdgeR workflow configuration
#     run_calcNormFactors = 'TMM'
#     estimateDisp_robust = FALSE
#     estimateDisp_trend.method = "locfit"
#     glm_approach = "QLF"
#     # Output configuration
#     adjust_method = 'BH'


```

### Functional enrichment
```{r gse-run, eval = FALSE}
library(org.Hs.eg.db)
dge_list <- readRDS(file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
dge <-  dge_list[[1]]
gse_list <- list()
gse_list <- foreach(dge = dge_list) %do%{
# for(dge_names in names(dge_list) ari) {
  # dge <- dge_list[[dge_names]]
  # cat('######', dge_names, "\n\n\n")
  cat('####', dge$comparison, "\n\n\n")
  # gse_list[[dge_names]] <- gse_omnibus(
  gse_res <- gse_omnibus(
    feature_names = dge$results$gene_name,
    p = dge$results$FDR,
    fc = dge$results$logFC,
    fc_thrs = 0.5,
    gmt_files = gmt_files_symbols,
    
    run_enricher = FALSE,
    run_enrichGO = FALSE,
    run_enrichKEGG = FALSE,
    run_enrichReactome = FALSE,
    
    run_GSEA = TRUE,
    run_gseGO = FALSE,
    run_gseKEGG = FALSE,
    run_gseReactome = FALSE,
    
    go_ontologies = c('BP'),
    args_gse = list(
      minGSSize = 10,
      maxGSSize = 250,
      pvalueCutoff = 1),
    args_enricher = list(
      minGSSize = 10,
      maxGSSize = 250, 
      pvalueCutoff = 1),
    go_simplify = TRUE,
    go_simplify_p = 0.1,
    go_simplify_cutoff = 0.6
    )
  
   return(gse_res)

}
names(gse_list) <- names(dge_list)

# Save objects
saveRDS(gse_list, file = file.path(params$output_dir,'gse_omnibus-edgeR_QLF.rds'))
```

### Load data
```{r load-deg-data}
dge_list <- readRDS(file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
gse_list <- readRDS(file.path(params$output_dir, 'gse_omnibus-edgeR_QLF.rds'))

# Annotate dge_list with dormancy genes
i <- names(dge_list)[[1]]
for (i in names(dge_list)) {
  dge_list[[i]]$results$is_dormancy <- 
    dge_list[[i]]$results$gene_name %in% dormancy_genes$gene_name
}
```



## Differential expression analysis
```{r knitr-template-rnaseq-deg-edger, eval = TRUE}
se$condition <- se$dormancy_brightness
group_palette <- get_scater_palette('tableau10medium')[1:nlevels(se$condition)] %>% 
  set_names(levels(se$condition))

# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-deg-edger.Rmd")
  )
```
`r knit(text = src)`

## Functional enrichment analysis
```{r knitr-template-rnaseq-gse-omnibus, eval = TRUE}
# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-gse-omnibus.Rmd")
  )
```
`r knit(text = src)`






## Cell cycle

```{r cc-phase-dormancy-barplot}
use_data <- colData(se) %>% data.frame %>% 
  group_by(dormancy_status, cc_phase) %>% 
  summarise(n = n()) %>% 
  ungroup()
use_data$p <- use_data %>% 
  reframe(
    p = n / sum(n), 
    .by = dormancy_status) %>% 
  pull(p)

use_data %<>%  arrange(dormancy_status, p)
use_data$pos <- use_data %>% 
  reframe(
    pos =  cumsum(p) - (0.5 * p), 
    .by = dormancy_status) %>% 
  pull(pos)

use_data %>% 
  ggplot(aes(dormancy_status, p, fill = cc_phase)) +
  geom_bar(stat="identity") +
  geom_text(aes(
    label = paste0(round(100*p, 1),"%"),
    y = pos), 
    size=geom_text_size) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_OkabeIto() +
  labs(
    x = '',
    y = 'Percentage',
    fill = ''
  )
```


```{r cc-phase-brightness-barplot}
use_data <- colData(se) %>% data.frame %>% 
  filter(dormancy_brightness %in% c('Bright', 'Dim', 'Cycling')) %>% 
  group_by(dormancy_brightness, cc_phase) %>% 
  summarise(n = n()) %>% 
  ungroup()
use_data$p <- use_data %>% 
  reframe(
    p = n / sum(n), 
    .by = dormancy_brightness) %>% 
  pull(p)

use_data %<>%  arrange(dormancy_brightness, p)
use_data$pos <- use_data %>% 
  reframe(
    pos =  cumsum(p) - (0.5 * p), 
    .by = dormancy_brightness) %>% 
  pull(pos)

use_data %>% 
  ggplot(aes(dormancy_brightness, p, fill = cc_phase)) +
  geom_bar(stat="identity") +
  geom_text(aes(
    label = paste0(round(100*p, 1),"%"),
    y = pos), 
    size=geom_text_size) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_OkabeIto() +
  labs(
    x = '',
    y = 'Percentage',
    fill = ''
  )
```
