---
title: "Recategorize BR16 dormant cells"
subtitle: "Recategorize cells based on differential expressed genes in Bright over Cycling"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  data_dir: ./output/rnaseq/br16_dtc
  dge_dir: ./output/rnaseq/br16_dtc/dge
  output_dir: ./output/rnaseq/br16_dtc/dge_hclusters
  se_file: se_GRCh38_RFP_mVenus_gene_qc.rds
  vector_dorm: mVenus
  min_counts: 3
  min_present_prop: 0.70
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 300, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)
library(dendextend)
library(dynamicTreeCut)
library(colorspace)
library(GSVA)
library(GSEABase)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/eigencorval.R')
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/subchunkify.R')
source('./code/R-functions/pairwise_terms_jc.r')
clean_msigdb_names <- function(x) x %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Define additional functions
```{r add-functions}
clean_sample_alias <- function(x) 
  gsub('o[0-9]*_[0-9]*-|_gDNA.*', '', x) %>% gsub(".*-MS", "MS", .) %>% gsub("_S[0-9]+_R[0-9].*", "", .) %>% gsub('__STARpass1', '', .) %>% gsub(".k2report", "", .) %>% gsub(".*-|.disambiguated.*", "", .)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Load SummarizedExperiment object
```{r load-se}
se <- readRDS(file.path(params$data_dir, paste0(params$se_file)))
```

Load list of dormancy genes
```{r load-dormancy-genes}
dormancy_genes <- read_csv(file = './data/resources/list_of_dormancy_genes_MS_20230823.csv')
```

Load differential expression and functional enrichment data
```{r load-deg-data}
dge_list <- readRDS(file.path(params$dge_dir, 'dge-edgeR_QLF.rds'))
gse_list <- readRDS(file.path(params$dge_dir, 'gse_omnibus-edgeR_QLF.rds'))

# Annotate dge_list with dormancy genes
i <- names(dge_list)[[1]]
for (i in names(dge_list)) {
  dge_list[[i]]$results$is_dormancy <- 
    dge_list[[i]]$results$gene_name %in% dormancy_genes$gene_name
}
```

Configure gene sets for functional enrichment
```{r conf-gse_omnbus}
msigdb_vs <- 'v2026.1.Hs'
gmt_files_symbols <- list(
  msigdb.h = paste0('./data/resources/MSigDB/', msigdb_vs, '/h.all.', msigdb_vs, '.symbols.gmt'),
  msigdb.c2.cp.reactome = paste0('./data/resources/MSigDB/', msigdb_vs, '/c2.cp.reactome.', msigdb_vs, '.symbols.gmt'),
  msigdb.c2.cp = paste0('./data/resources/MSigDB/', msigdb_vs, '/c2.cp.', msigdb_vs, '.symbols.gmt')
)
```

## Data generation

### Data wrangling

Modify sample annotation
```{r dw-coldata, eval = FALSE}
colData(se) <- colData(se) %>% data.frame %>% 
  mutate(
    ## dormancy_brightness
    dormancy_brightness = factor(dormancy_brightness,
                                 levels = c('Bright',
                                            'Mixed_dim_bright',
                                            'Dim', 
                                            'Mixed_dim_cycling', 
                                            'Cycling', 
                                            'Unk')),
    dormancy_brightness = droplevels(dormancy_brightness),
    ## check dormancy status
    dormancy_check = ifelse(dormancy_check == 'no',
                            'Regular',
                            dormancy_check),
    dormancy_check = ifelse(grepl('Picked_on_G_Floor_CC', dormancy_check),
                            'Different microscope',
                            dormancy_check),
    dormancy_check = factor(dormancy_check) %>% relevel(ref = 'Regular'),
    ## sample_type
    sample_type_o=sample_type,
    sample_type = gsub(";wbc|;dn", "", sample_type),
    sample_type = factor(sample_type) %>% relevel(ref = 'dtc_single'),
    #time_months
    time_months = factor(time_months, levels = as.numeric(time_months) %>% unique %>% sort %>% as.character)
  ) %>% 
  DataFrame
```

Select only Bright, Dim and Cycling samples
```{r dw-select-samples, eval = FALSE}
se <- se[,se$dormancy_brightness %in% c('Bright', 'Dim', 'Cycling')]
se$dormancy_brightness <- droplevels(se$dormancy_brightness)
```

Filtering out low-abundance features. Keeping those features with at least 10 counts in at least 70% of samples of the smallest group.
```{r dw-subset-feature-abundance, eval = FALSE}
use_rows <- filterByExpr(se, group=paste(se$sample_type, se$dormancy_brightness), min.count = 1, min.prop = 0.3)
se <- se[use_rows,]

# Update PerCellQCMetrics
colData(se) <- colData(se) %>% data.frame %>% dplyr::select(-(sum:subsets_Ribo_percent)) %>% DataFrame
se <- addPerCellQCMetrics(
  se,
  subsets=list(
    Mito=which(rowData(se)$is.mito),
    Ribo=which(rowData(se)$is.ribo)
  ),
  percent.top = c(1,5,10,20, 50, 100),
  threshold = 3
  )

# # Table of removed features
# table(use_rows) %>% 
#   data.frame %>% 
#   mutate(use_rows = ifelse(use_rows == TRUE, 'keep', 'remove')) %>% 
#   column_to_rownames('use_rows') %>% 
#   kbl(caption = 'Filtered features') %>%
#   kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

Gene expression normalization
```{r dw-tmm, eval = FALSE}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)

assay(se, 'tpm') <- calculateTPM(se, assay.type = 'normcounts', rowData(se)$Length)

assay(se, 'cpm') <- edgeR::cpm(assay(se, 'normcounts'))
assay(se, 'logcpm') <- edgeR::cpm(assay(se, 'normcounts'), log = TRUE)
```

Cell cycle annotation with cyclone
```{r dw-cc-cyclone, eval = FALSE}
cc.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", 
    package="scran"))
res <- cyclone(se, cc.pairs)
se$cc_phase <- res$phases
se$cc_g1_score <- res$normalized.scores$G1
se$cc_s_score <- res$normalized.scores$S
se$cc_g2m_score <- res$normalized.scores$G2M
```

Run [GSVA](https://www.bioconductor.org/packages/release/bioc/vignettes/GSVA/inst/doc/GSVA.html) with dormancy genes manually selected.
```{r dw-gsva-dormancy-manual-genes, eval = FALSE}
use_assay <- assay(se, 'logcpm')
rownames(use_assay) <- make.names(rowData(se)$gene_name, unique = TRUE)

use_gsets <- list(
  `dormancy_ms_up` = dormancy_genes %>% filter(Direction == 'Up') %>% pull(gene_name) %>% unique,
  `dormancy_ms_down` = dormancy_genes %>% filter(Direction == 'Down') %>% pull(gene_name) %>% unique
)

# Generate GSVA scores
gsva_params <- gsvaParam(use_assay,
                         use_gsets,
                         kcdf="Gaussian",
                         maxDiff = TRUE)

res <- gsva(gsva_params)

res <- res[, colnames(se)] %>% t()
for(i in colnames(res)) {
  colData(se)[,i]<- res[,i]
}
```

Run [GSVA](https://www.bioconductor.org/packages/release/bioc/vignettes/GSVA/inst/doc/GSVA.html)  with MSigDb with p27 gene and related to senescence and negative regulation of cell cycle
```{r dw-gsva-dormancy, eval = FALSE}
use_assay <- assay(se, 'logcpm')
rownames(use_assay) <- make.names(rowData(se)$gene_name, unique = TRUE)

gset_c2 <- getGmt('./data/resources/MSigDB/v2026.1.Hs/c2.cp.v2026.1.Hs.symbols.gmt')
gset_c5 <- getGmt('./data/resources/MSigDB/v2026.1.Hs/c5.go.bp.v2026.1.Hs.symbols.gmt')
gset_c2cgp <- getGmt('./data/resources/MSigDB/v2026.1.Hs/c2.cgp.v2026.1.Hs.symbols.gmt')

use_gsets <- list(
  TOP_DEG_DOWN = c('CDC20','PSRC1','KIF20A','ASPM','HJURP','PTTG1','DLGAP5','TTK','FAM83D','AURKB','CENPE','CDCA3','ARHGAP11B','KNL1','PLK1','TROAP','SHCBP1','NUF2','TRAIP','KIF2C'),
  BIOCARTA_P27_PATHWAY = geneIds(gset_c2[['BIOCARTA_P27_PATHWAY']]),
  REACTOME_G0_AND_EARLY_G1 = geneIds(gset_c2[['REACTOME_G0_AND_EARLY_G1']]),
  REACTOME_CELLULAR_SENESCENCE = geneIds(gset_c2[['REACTOME_CELLULAR_SENESCENCE']]),
  REACTOME_MITOTIC_G1_PHASE_AND_G1_S_TRANSITION = geneIds(gset_c2[['REACTOME_MITOTIC_G1_PHASE_AND_G1_S_TRANSITION']]),
  REACTOME_G1_S_SPECIFIC_TRANSCRIPTION = geneIds(gset_c2[['REACTOME_G1_S_SPECIFIC_TRANSCRIPTION']]),
  GOBP_NEGATIVE_REGULATION_OF_MITOTIC_CELL_CYCLE = geneIds(gset_c5[['GOBP_NEGATIVE_REGULATION_OF_MITOTIC_CELL_CYCLE']]),
  GOBP_NEGATIVE_REGULATION_OF_MITOTIC_CELL_CYCLE_PHASE_TRANSITION = geneIds(gset_c5[['GOBP_NEGATIVE_REGULATION_OF_MITOTIC_CELL_CYCLE_PHASE_TRANSITION']]),
  GOBP_CELLULAR_SENESCENCE = geneIds(gset_c5[['GOBP_CELLULAR_SENESCENCE']]),
  FRIDMAN_SENESCENCE_DN = geneIds(gset_c2cgp[['FRIDMAN_SENESCENCE_DN']]),
  FRIDMAN_SENESCENCE_UP = geneIds(gset_c2cgp[['FRIDMAN_SENESCENCE_UP']]),
  HVG_G1_S_TRANSCRIPTION = c('RRM2','CDK1','CDT1','CDC6','TK1','TYMS','CCNE1','E2F1','CDC45'),
  `CA2+_Binding_proteins` = c('S100P', 'S100A14', 'S100A6', 'S100A16'),
  `Genes_related_to_CA2+` = c('PSEN2', 'CRABP2', 'GSN', 'ANXA2', 'ANXA6', 'CIB1')
)

## ADD G0 from GO
go_g0_names <- grep('G0', names(gset_c5), value = TRUE)
for(i in go_g0_names){
  use_gsets[[i]] <- geneIds(gset_c5[[i]])
}

# Generate GSVA scores
gsva_params <- gsvaParam(use_assay,
                         use_gsets,
                         kcdf="Gaussian",
                         maxDiff = TRUE)

res <- gsva(gsva_params)

res <- res[, colnames(se)] %>% t()
for(i in colnames(res)) {
  colData(se)[,i]<- res[,i]
}

```


Save final SummarizedExperiment object after data wrangling and normalization
```{r dw-save-se, eval = FALSE}
saveRDS(se, file = file.path(params$output_dir, 'se.rds'))
```

Load SummarizedExperiment object after data wrangling and normalization
```{r dw-load-se}
se <- readRDS(file.path(params$output_dir, 'se.rds'))
```

### Hyerarchical clustering

```{r recat-dge-brightness-config}
p_thrs <- 0.001
lfc_thrs <- 1
i <- "bright-o-cycling"

# Extract DGE results
use_dge <- dge_list[[i]]$results %>% 
  filter(FDR <= p_thrs & abs(logFC) >= lfc_thrs) %>% 
  mutate(direction = ifelse(logFC > 0, 'Up', 'Down')) %>% 
  arrange(FDR) %>% 
  arrange(logFC)

```

Select genes for the analysis and generate matrix for clustering and exploratory analysis. We use the genes that are differentially expressed in Bright over Cycling samples, with FDR ≤ `r p_thrs`  and absolute log2 fold-change ≥ `r lfc_thrs` (n genes = `r nrow(use_dge)`). We scale the expression of these genes across samples to generate a matrix for clustering and exploratory analysis.

```{r recat-dge-brightness-zmat}
mat <- assay(se, 'logcounts')[rownames(use_dge), ]
rownames(mat) <- make.names(use_dge$gene_name, unique = TRUE)
zmat <- t(apply(mat, 1, scale, center = TRUE, scale = TRUE))
colnames(zmat) <- colnames(mat)
```


Perform sample clustering and dendrogram. We cut the dendrogram to obtain the clusters.
```{r recat-dge-brightness-hclust-dend, eval = FALSE}
# Perform clustering and dendrogram
dst <- t(zmat) %>% dist(method = 'euclidean')
hc <-  dst %>% hclust(method = 'ward.D2')
dend <- hc %>% as.dendrogram()
column_dend<- dend


## Find Clusters
# clusters_original <- cutreeDynamic(hc, distM = as.matrix(dst),
#                                    minClusterSize = 3, method="hybrid")
clusters_original <- cutree(hc, k = 5)

## Add clustering info to se
# cluster_info_df <- data.frame(
#   sample_alias = colnames(zmat),
#   cluster = as.character(clusters_original) %>%
#     factor(levels = c('4', '1', '5', '2', '3')),
#   cluster_cycling = clusters_original %in% c('4', '1'),
#   cluster_dormant = clusters_original %in% c('5', '2', '3')
#   # cluster = factor(as.character(clusters_original))
#   )

cluster_info_df <- data.frame(
  sample_alias = colnames(zmat),
  # cluster = as.character(clusters_original),
  cluster = case_when(
    clusters_original == '1' ~ 'mVenus neg 1',
    clusters_original == '2' ~ 'mVenus pos 2',
    clusters_original == '3' ~ 'mVenus pos 1',
    clusters_original == '4' ~ 'mVenus neg 2',
    clusters_original == '5' ~ 'mVenus pos 3'
  ),
  cluster_cycling = clusters_original %in% c('4', '1'),
  cluster_dormant = clusters_original %in% c('5', '2', '3')
  # cluster = factor(as.character(clusters_original))
  )

```

Save cluster information
```{r recat-dge-brightness-hclust-save, eval = FALSE}
cluster_info <- list(
  cluster_info_df = cluster_info_df,
  clusters_original = clusters_original,
  dst = dst,
  hc = hc,
  dend = dend,
  column_dend = column_dend 
)
saveRDS(cluster_info, file = file.path(params$output_dir, 'cluster_info.rds'))
```

Add cluster info to summarized experiment and generate colored dendrogram. 
```{r recat-dge-brightness-hclust-to-se}
cluster_info <- readRDS(file.path(params$output_dir, 'cluster_info.rds')) 
if(!is.null(se$cluster))
  se$cluster <- NULL
colData(se) <- colData(se) %>% data.frame %>% 
  left_join(cluster_info$cluster_info_df) %>% 
  mutate(torownames = sample_alias) %>%
  column_to_rownames('torownames') %>% 
  DataFrame


## Sort clusters to the order of the dendrogram:
clusters <- cluster_info$clusters_original[order.dendrogram(cluster_info$dend)]
clusters_numbers <- unique(clusters) - (0 %in% clusters)
n_clusters <- length(clusters_numbers)
cols <- rainbow_hcl(n_clusters)
column_dend_colored <- cluster_info$dend %>% 
         branches_attr_by_clusters(clusters, values = cols)
```

### Differential expression and functional enrichment
```{r conf-comp}
# Coldata
x <- colData(se) %>% data.frame 

# # Coldata for Bright and Dim samples only
# xbd <- colData(se) %>% data.frame %>% 
#   filter(dormancy_brightness %in% c('Bright', 'Dim'))

comp_list <- list(
  `Dormant--over--Cycling` = list(
    Dormant = x %>% 
      filter(cluster_dormant) %>% 
      pull(sample_alias) %>% 
      sort,
    Cycling = x %>% 
      filter(cluster_cycling) %>% 
      pull(sample_alias) %>% 
      sort
  ),
  `mVenus_pos_3--over--1-2` = list(
    mVenus_pos_3 = x %>% 
      filter(cluster == 'mVenus pos 3') %>% 
      pull(sample_alias) %>% 
      sort,
    mVenus_pos_1_2 = x %>% 
      filter(cluster %in% c('mVenus pos 1', 'mVenus pos 2')) %>% 
      pull(sample_alias) %>% 
      sort
  ),
  `mVenus_pos_2--over--1-3` = list(
    mVenus_pos_2 = x %>% 
      filter(cluster == 'mVenus pos 2') %>% 
      pull(sample_alias) %>% 
      sort,
    mVenus_pos_1_3 = x %>% 
      filter(cluster %in% c('mVenus pos 1', 'mVenus pos 3')) %>% 
      pull(sample_alias) %>% 
      sort
  ),
  `mVenus_pos_1--over--2-3` = list(
    mVenus_pos_1 = x %>% 
      filter(cluster == 'mVenus pos 1') %>% 
      pull(sample_alias) %>% 
      sort,
    mVenus_pos_2_3 = x %>% 
      filter(cluster %in% c('mVenus pos 2', 'mVenus pos 3')) %>% 
      pull(sample_alias) %>% 
      sort
  )
)


comp_list_description <- data.frame(
  comparison = names(comp_list),
  comparison_long = c(
    'Dormant over Cycling - cluster based',
    'Cluster mVenus pos 3 over 1+2',
    'Cluster mVenus pos 2 over 1+3',
    'Cluster mVenus pos 1 over 2+3'
  )
)
```

Run differential expression analysis using EdgeR QLF
```{r run-dge, eval = FALSE}
ensembl <- biomaRt::useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
gene_desc <- biomaRt::getBM(attributes=c('external_gene_name','description'),
                              filters = 'external_gene_name',
                              values = rowData(se)$gene_name, mart =ensembl) %>%
    dplyr::rename('gene_name' = 'external_gene_name') %>%
    unique

use_comp <-  comp_list[[1]]
dge_list <- foreach(use_comp = comp_list) %do%{
  se_cols <- se$sample_alias %in% unlist(use_comp)
  use_se <- se[,se_cols]
  use_se$group <- ifelse(
    use_se$sample_alias %in% use_comp[[1]],
    names(use_comp)[1],
    names(use_comp)[2]
  )
  
  if(is.null(use_comp$batch)) {
    dge <- edgeR_dge(
      use_se,
      # Design configuration for differential expression
      group_var =  'group',
      group_sample = names(use_comp)[1],
      group_ref = names(use_comp)[2],
      design_formula = "~ group",
      coef = 'last',
      # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
      assay_to_DGEList = 'counts',
      assay_to_row_filter = "counts",
      # Feature filtering parameters
      use_filterByExpr = TRUE,
      min_counts = params$min_counts,
      min_present_prop = params$min_present_prop,
      # EdgeR workflow configuration
      run_calcNormFactors = 'TMM',
      estimateDisp_robust = TRUE,
      estimateDisp_trend.method = "locfit",
      glm_approach = "QLF",
      # Output configuration
      adjust_method = 'BH'
      )
    } else if(use_comp$batch == 'donor') {
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        batch_vars = 'donor',
        design_formula = "~ donor + group",
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
        )
    }

  use_res <- dge$results %>%  left_join(., gene_desc)
  dge$results <- use_res %>% 
    filter(!duplicated(feature)) %>% 
    mutate(rownames = feature) %>% 
    column_to_rownames('rownames')
  
   return(dge)
}

names(dge_list) <- names(comp_list)
detach("package:biomaRt", unload=TRUE)
# Save objects
saveRDS(dge_list, file = file.path(params$output_dir, 'dge-edgeR_QLF.rds'))

###########
#   use_samples = NULL
#   # Desing configuration for differential expression
#   group_var = NULL
#   group_sample = NULL
#   group_ref = NULL
#   numeric_covar = NULL
#   batch_vars = NULL
#   design_formula = NULL
#   contrast = NULL
#   coef = NULL
#   # Conversion from SingleCellExpeiment to DGEList
#   spike_normalization = FALSE
#   assay_to_DGEList = 'counts'
#   assay_to_row_filter = "tpm"
#   use_colData = NULL
#   use_rowData = NULL
#   # Feature filtering parameters
#   use_filterByExpr = TRUE
#   min_counts = 1
#   min_present_prop = 0.40
#   # EdgeR workflow configuration
#   run_calcNormFactors = 'TMM'
#   estimateDisp_robust = FALSE
#   estimateDisp_trend.method = "locfit"
#   glmQLFit_robust = FALSE
#   glm_approach = "QLF"
#   # Output configuration
#   adjust_method = 'BH'
#   assays_from_SingleCellExperiment = NULL
# 
# object = se
#     # Desing configuration for differential expression
#     group_var =  condition_var
#     group_sample = use_comp[1]
#     group_ref = use_comp[2]
#     batch_vars = 'sample_type'
#     design_formula = paste("~ 0 +", condition_var," + sample_type")
#     coef = NULL
#     # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
#     assay_to_DGEList = 'counts'
#     assay_to_row_filter = "counts"
#     # Feature filtering parameters
#     use_filterByExpr = TRUE
#     min_counts = params$min_counts
#     min_present_prop = params$min_present_prop
#     # EdgeR workflow configuration
#     run_calcNormFactors = 'TMM'
#     estimateDisp_robust = FALSE
#     estimateDisp_trend.method = "locfit"
#     glm_approach = "QLF"
#     # Output configuration
#     adjust_method = 'BH'


```

Run functional enrichment
```{r run-gse, eval = FALSE}
library(org.Hs.eg.db)
dge_list <- readRDS(file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
dge <-  dge_list[[1]]
gse_list <- list()
gse_list <- foreach(dge = dge_list) %do%{
# for(dge_names in names(dge_list) ari) {
  # dge <- dge_list[[dge_names]]
  # cat('######', dge_names, "\n\n\n")
  cat('####', dge$comparison, "\n\n\n")
  # gse_list[[dge_names]] <- gse_omnibus(
  gse_res <- gse_omnibus(
    feature_names = dge$results$gene_name,
    p = dge$results$FDR,
    fc = dge$results$logFC,
    fc_thrs = 0.5,
    gmt_files = gmt_files_symbols,
    
    run_enricher = TRUE,
    run_enrichGO = TRUE,
    run_enrichKEGG = FALSE,
    run_enrichReactome = TRUE,
    
    run_GSEA = TRUE,
    run_gseGO = TRUE,
    run_gseKEGG = FALSE,
    run_gseReactome = TRUE,
    
    go_ontologies = c('BP'),
    args_gse = list(
      minGSSize = 10,
      maxGSSize = 250,
      pvalueCutoff = 1),
    args_enricher = list(
      minGSSize = 10,
      maxGSSize = 250, 
      pvalueCutoff = 1),
    go_simplify = TRUE,
    go_simplify_p = 0.1,
    go_simplify_cutoff = 0.6
    )
  
   return(gse_res)

}
names(gse_list) <- names(dge_list)

# Save objects
saveRDS(gse_list, file = file.path(params$output_dir,'gse_omnibus-edgeR_QLF.rds'))
```

Load differential expression and functional enrichment results
```{r load-dge-gse}
dge_list <- readRDS(file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
gse_list <- readRDS(file.path(params$output_dir, 'gse_omnibus-edgeR_QLF.rds'))
```

## Sample clustering
```{r recat-dge-brightness-hclust-dend-plot, fig.width=7.2}
cluster_info <- readRDS(file.path(params$output_dir, 'cluster_info.rds')) 


## Sort clusters to the order of the dendrogram:
clusters <- cluster_info$clusters_original[order.dendrogram(cluster_info$dend)]
clusters_numbers <- unique(clusters) - (0 %in% clusters)
n_clusters <- length(clusters_numbers)
# cols <- rainbow_hcl(n_clusters)
cols <- cluster_palette
column_dend_colored <- cluster_info$dend %>% 
         branches_attr_by_clusters(clusters, values = cols)

## Define column annotation colors
use_annot_var <- se$dormancy_brightness %>%
  as.character
annot_color <- palette_dormancy_brightness[use_annot_var]

# Gnerate cluster palette for heatmap column annotation
# cluster_palette <- cols %>% set_names(unique(clusters))
# Plot colored dendrogram
plot(column_dend_colored)
colored_bars(colors = annot_color, dend = column_dend_colored, rowLabels = "Brightness")
```

## PCA
```{r recat-dge-brightness-pca-conf}
ann_df <- colData(se) %>% data.frame %>% 
  dplyr::select(
    detected,
    sum,
    dormancy_check,
    dormancy_brightness,
    centrifugation_g,
    time_months,
    anatomical_origin,
    sample_type,
    dormancy_status,
    mouse_id,
    cluster)

mat <- assay(se, 'logcounts')[rownames(use_dge), ]

p <- PCAtools::pca(mat, 
                   metadata = ann_df[colnames(mat),]
                   )
```

#### Variable correlation
Plot showing the correlation of PCs with sample-level variables. 
```{r recat-dge-brightness-pca-eigencor-plot, fig.width = 7.2}
n_pc <- 1:5

eigencor_res <- eigencorval(p,components = getComponents(p, n_pc), metavars = colnames(p$metadata))

# Select variables correlated with PC1 and PC2
col_vars <- colnames(eigencor_res$pvals)
col_pc12 <- c(
  col_vars[eigencor_res$pvals['PC1',]<0.05],
  col_vars[eigencor_res$pvals['PC2',]<0.05]
)

# Plot correlation
eigencorplot(p,
             components = getComponents(p, n_pc),
             metavars = colnames(p$metadata),
             # col = c('darkblue', 'blue2', 'black', 'red2', 'darkred'),
             # col = c('gold', 'cornsilk1', 'white', 'forestgreen', 'darkgreen'),
             # colCorval = 'white',
             fontCorval = 2,
             main = 'PC sample variables correlation')


```

#### Bi-plot
```{r recat-dge-brightness-pca-pc1-pc2-plot, fig.asp=0.8, fig.width = 7.2}
i <- 'dormancy_brightness'
PCAtools::biplot(
  p, 
  colby = i, 
  colkey = palette_dormancy_brightness,
  colLegendTitle = i,
  shape = 'sample_type',
  shapeLegendTitle = 'Sample Type',
  # encircle = use_encircle,
  # encircleAlpha = 1/4,
  legendPosition = 'right',
  pointSize = 3,
  lab = NULL
  # loadings parameters
  # showLoadings = TRUE,
  # lengthLoadingsArrowsFactor = 1.5,
  # elipse config
  # ellipse = use_ellipse,
  # ellipseType = 't',
  # ellipseLevel = 0.75
) +
  theme_facet
```

```{r recat-dge-brightness-pca-pc1-pc2-plot-hclust, fig.asp=0.8, fig.width = 3.2}
i <- 'cluster'
PCAtools::biplot(
  p, 
  colby = i, 
  colkey = cluster_palette,
  colLegendTitle = i,
  
  shape = 'dormancy_brightness',
  shapeLegendTitle = 'Fluorescence',
  # encircle = use_encircle,
  # encircleAlpha = 1/4,
  legendPosition = 'right',
  pointSize = 1,
  lab = NULL, 

  # elipse config
  ellipse = TRUE,
  ellipseType = 't',
  ellipseLevel = 0.90
) +
  theme_project



```

```{r recat-dge-brightness-pca-pc1-pc2-plot-hclust-ggplot, fig.asp=0.8, fig.width = 3.2}
cbind(p$metadata, p$rotated[,1:2]) %>% 
  ggplot(aes(x = PC1, y = PC2, color = cluster, shape = dormancy_brightness)) +
  geom_point(size = 2, alpha = 0.75) +
  scale_color_manual(values = cluster_palette) +
  
  labs(
    shape = 'Fluorescence',
    color = 'Cluster',
    x = paste("PC\n", round(p$var[1], 2), "% variation", sep = ""),
    y = paste("PC\n", round(p$var[2], 2), "% variation", sep = "")
  )


cbind(p$metadata, p$rotated[,1:2]) %>% 
  ggplot(aes(x = PC1, y = PC2, color = time_months, shape = dormancy_brightness)) +
  geom_point(size = 5, alpha = 1) +
  # scale_color_manual(values = cluster_palette) +
  
  labs(
    shape = 'Fluorescence',
    color = 'Cluster',
    x = paste("PC\n", round(p$var[1], 2), "% variation", sep = ""),
    y = paste("PC\n", round(p$var[2], 2), "% variation", sep = "")
  )

```

## Heatmap unsupervised clustering
Generate heatmap with pre-defined dendrogram and clustering
```{r recat-dge-brightness-heatmap-hclust-heatmap, fig.width=7.8, fig.asp = 1.6}
max_color_range <- 3

# Column Annotation
use_annot <-  colData(se) %>% data.frame %>% 
  dplyr::select(dormancy_brightness, 
                cluster
                # sample_type,
                # ctc_n,
                # time_months,
                # anatomical_origin,
                # centrifugation_g
                ) %>% 
  dplyr::select(cluster, everything()) %>% 
  dplyr::rename(
    Condition = dormancy_brightness,
    Cluster = cluster
    # `DTC type` = sample_type,
    # `DTC cells` = ctc_n,
    # `Time (m)` = time_months,
    # `Site` = anatomical_origin,
    # `Centrifugation (g)` = centrifugation_g
  )
use_annot <- use_annot[colnames(zmat),]
  
palette_ctc_n <- colorRamp2(
  c(1, max(use_annot$`DTC cells`)), 
  c("#ccece6", "#006d2c")
  )
column_ha <-  HeatmapAnnotation(
  df = use_annot,
  col = list(
    Condition = palette_dormancy_brightness[levels(use_annot$Condition)],
    Cluster = cluster_palette
    # `DTC type` = palette_sample_type,
    # `Time (m)` = palette_time,
    # `Site` = palette_anatomy,
    # `Centrifugation (g)` = palette_centrifucation
  ),
  annotation_name_gp = gpar(fontsize = 8),
  annotation_name_side = 'left'
)


# Row annotation (right)
row_annot_right <-  rowAnnotation(
  log2FC = anno_numeric(
    round(use_dge$logFC,1), 
    labels_gp = gpar(fontsize = 6),
    bg_gp = gpar(
      fill = c("#74add1", "#f46d43"), 
      col = 'white'
    ),
    bar_width = unit(1, "npc") - unit(1.5, "pt")
  ), 
  annotation_name_rot = 0)


# Row annotation (left)
direction_palette <- c('Up' = "#f46d43", 'Down' = "#74add1")
direction_palette <- direction_palette[names(direction_palette) %in% use_dge$direction]

row_annot_left <-  rowAnnotation(
  direction = anno_block(
    gp = gpar(fill = direction_palette),
    labels = names(direction_palette), 
    labels_gp = gpar(col = "white", fontsize = 8)
  )
)


# Heatmap color
zmax <- zmat %>% abs %>% max
if(!is.null(max_color_range))
  zmax <- max_color_range
col_fun <-  colorRamp2(
  seq(-zmax, zmax, length.out = 11),
  rev(brewer.pal(11, "RdBu"))
)

# Generate Heatmap
htz <- Heatmap(
  zmat,
  name= 'z-score',
  col = col_fun,
  
  row_names_gp = gpar(fontsize = 6),
  # column_names_gp = gpar(fontsize = 8),
  show_column_names = FALSE,
  
  top_annotation = column_ha,
  cluster_columns = column_dend_colored,
  column_title = NULL,
  
  show_row_dend = FALSE,
  column_split = 2,

  row_title = NULL
)


draw(htz, 
     annotation_legend_side = "bottom"
)
cat("\n\n")
```

Generate heatmap split by sorted clusters
```{r recat-dge-brightness-heatmap-hclust-sorted-heatmap, fig.width=7.8, fig.asp = 1.6}
# Generate Heatmap
htz <- Heatmap(
  zmat,
  name= 'z-score',
  col = col_fun,
  row_names_gp = gpar(fontsize = 6),
  # column_names_gp = gpar(fontsize = 8),
  show_column_names = FALSE,
  
  top_annotation = column_ha,
  # cluster_columns = column_dend_colored,
  clustering_method_columns = 'ward.D2',
  column_split = factor(use_annot$Cluster, levels = c("mVenus neg 2", "mVenus neg 1", "mVenus pos 3", "mVenus pos 2", "mVenus pos 1")),
  cluster_column_slices = FALSE,
  column_title = NULL,
  
  
  show_row_dend = FALSE,

  row_title = NULL
)


draw(htz, 
     # heatmap_legend_side = "bottom", 
     annotation_legend_side = "bottom"
)
cat("\n\n")
```


## Clusters analysis
```{r recatdge-brightness-cluster-analysis}
my_comparisons <- list(
  c("1", "4"),
  c("1", "3"),
  c("1", "5"),
  c("1", "2")
)
    
```

### Vector expression by cluster
```{r recat-dge-brightness-vector-exp}
use_data <- colData(se) %>% data.frame
use_data$vector <- assay(se, 'logcounts')[params$vector_dorm,]

use_data %>% 
  ggplot(aes(cluster, vector, fill = cluster)) +
  geom_violin(width=0.9, scale = 'width', linewidth = rel(0.4)) +
  stat_summary(data = use_data %>% filter(dormancy_status!='mixed_dormant'),
               fun.data="median_q1q3", 
               mult=1, 
               geom="crossbar", 
               width=0.08, 
               color = alpha('black', alpha = 1),
               fill = 'white'
               ) +
  # geom_jitter(shape=16, width=0.1, alpha = 0.8, show.legend = FALSE) +
  # geom_dotplot(shape=16, binaxis='y', stackdir='center', dotsize=0.9, alpha = 0.8) +
  geom_quasirandom(shape = 16, alpha = 0.5, width=0.25, size = rel(0.8)) +
  scale_fill_manual(values = cluster_palette) +
  labs(
    x = 'Clusters',
    y = paste(params$vector_dorm, "expression"),
    fill = '',
    caption = paste(
      'Y-axis: normalized expression in log scale',
      'Points represent individual observations',
      'Box represent Q25, median and Q75',
      sep = '\n'
    )
  ) + 
  guides(fill = 'none') +
  # theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  stat_compare_means(
    comparisons = my_comparisons,
    label.y.npc = 1,
    label.x.npc = 'left',
    size = 0.8*geom_text_size
    )
  cat("\n\n")

```

### Phenotype and quality variables
```{r recat-dge-brightness-pheno-exp}
chunk_data <- colData(se) %>% data.frame %>% 
  mutate(
    time_months = as.character(time_months) %>% as.numeric,
    `log (N reads)` = log(sum)
    )

test_var <- c('log (N reads)', 'detected', 
              'time_months', 'ctc_n',
              'subsets_Mito_percent', 
              'subsets_Mito_detected',
              'subsets_Ribo_percent', 
              'subsets_Ribo_detected'
              ) 

for (i in test_var) {
  res <- chunk_data %>% 
    ggplot(aes(cluster, .data[[i]], fill = cluster)) +
    geom_violin(width=0.9, scale = 'width', linewidth = rel(0.4)) +
    stat_summary(data = chunk_data %>% filter(!dormancy_status %in% c('Mixed', 'Unk')),
                 fun.data="median_q1q3", 
                 mult=1, 
                 geom="crossbar", 
                 width=0.08, 
                 color = alpha('black', alpha = 1),
                 fill = 'white'
                 ) +
    geom_quasirandom(shape = 16, alpha = 0.5, width=0.25, size = rel(0.8)) +
    scale_fill_manual(values = cluster_palette) +
    labs(
      title = i,
      x = 'Clusters',
      y = i,
      fill = '',
      caption = paste(
        'Y-axis: normalized expression in log scale',
        'Points represent individual observations',
        'Box represent Q25, median and Q75',
        sep = '\n'
      )
    ) + 
    guides(fill = 'none') +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) +
    stat_compare_means(
      comparisons = my_comparisons,
      label.y.npc = 1,
      label.x.npc = 'left',
      size = 0.8*geom_text_size
      )
  print(res)
  cat("\n\n")
}
```

### Cell cycle scores
```{r recat-dge-brightness-ccs-exp}
chunk_data <- colData(se) %>% data.frame

test_var <- c('cc_g1_score', 'cc_s_score', 'cc_g2m_score') 

for (i in test_var) {
  res <- chunk_data %>% 
    ggplot(aes(cluster, .data[[i]], fill = cluster)) +
    geom_violin(width=0.9, scale = 'width', linewidth = rel(0.4)) +
    stat_summary(data = chunk_data %>% filter(!dormancy_status %in% c('Mixed', 'Unk')),
                 fun.data="median_q1q3", 
                 mult=1, 
                 geom="crossbar", 
                 width=0.08, 
                 color = alpha('black', alpha = 1),
                 fill = 'white'
                 ) +
    geom_quasirandom(shape = 16, alpha = 0.5, width=0.25, size = rel(0.8)) +
    scale_fill_manual(values = cluster_palette) +
    labs(
      title = gsub("_score", "", i) %>% gsub("cc_", "", .) %>% toupper,
      x = 'Clusters',
      y = i,
      fill = '',
      caption = paste(
        'Y-axis: normalized expression in log scale',
        'Points represent individual observations',
        'Box represent Q25, median and Q75',
        sep = '\n'
      )
    ) + 
    guides(fill = 'none') +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) +
    stat_compare_means(
      comparisons = my_comparisons,
      label.y.npc = 1,
      label.x.npc = 'left',
      size = 0.8*geom_text_size
      )
  print(res)
  cat("\n\n")
}
```

```{r cc-score-dormancy-scatterplot}
colData(se) %>% data.frame %>% 
  ggplot(aes(cc_g1_score, cc_g2m_score, color = cluster)) +
  geom_point(shape = 16, size = rel(2), alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = 2) +
  geom_vline(xintercept = 0.5, linetype = 2) +
  scale_color_manual(values =  cluster_palette) +
  labs(
    x = 'G1 score',
    y = 'G2M score',
    color = '',
    caption = 'Cells are classified as being in G1 phase if the G1 score is above 0.5 and greater than the G2/M score; in G2/M phase if the G2/M score is above 0.5 and greater than the G1 score; and in S phase if neither score is above 0.5.'
  )

```

### Cell cycle phase
```{r cc-phase-brightness-barplot}
chunk_data <- colData(se) %>% data.frame %>% 
  group_by(cluster, cc_phase) %>% 
  summarise(n = n()) %>% 
  ungroup()
chunk_data$p <- chunk_data %>% 
  reframe(
    p = n / sum(n), 
    .by = cluster) %>% 
  pull(p)

chunk_data %<>%  arrange(cluster, p)
chunk_data$pos <- chunk_data %>% 
  reframe(
    pos =  cumsum(p) - (0.5 * p), 
    .by = cluster) %>% 
  pull(pos)

chunk_data %>% 
  ggplot(aes(cluster, p, fill = cc_phase)) +
  geom_bar(stat="identity") +
  geom_text(aes(
    label = paste0(round(100*p, 1),"%"),
    y = pos), 
    size=geom_text_size) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_OkabeIto() +
  labs(
    x = '',
    y = 'Percentage',
    fill = ''
  )
```

### Dormancy gene sets
```{r recat-dge-brightness-gsva-exp}
chunk_data <- colData(se) %>% data.frame
test_var <- chunk_data %>% 
  dplyr::select(dormancy_ms_up:GOBP_POSITIVE_REGULATION_OF_G0_TO_G1_TRANSITION) %>% 
  colnames()

for (i in test_var) {
  res <- chunk_data %>% 
    ggplot(aes(cluster, .data[[i]], fill = cluster)) +
    geom_violin(width=0.9, scale = 'width', linewidth = rel(0.4)) +
    stat_summary(data = chunk_data %>% filter(!dormancy_status %in% c('Mixed', 'Unk')),
                 fun.data="median_q1q3", 
                 mult=1, 
                 geom="crossbar", 
                 width=0.08, 
                 color = alpha('black', alpha = 1),
                 fill = 'white'
                 ) +
    geom_quasirandom(shape = 16, alpha = 0.5, width=0.25, size = rel(0.8)) +
    scale_fill_manual(values = cluster_palette) +
    labs(
      title = i,
      x = 'Clusters',
      y = 'GSVA score',
      fill = '',
      caption = paste(
        'Y-axis: normalized expression in log scale',
        'Points represent individual observations',
        'Box represent Q25, median and Q75',
        sep = '\n'
      )
    ) + 
    guides(fill = 'none') +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) +
    theme(
      plot.title = element_text(size = 6),
    ) +
    stat_compare_means(
      comparisons = my_comparisons,
      label.y.npc = 1,
      label.x.npc = 'left',
      size = 0.8*geom_text_size
      )
  print(res)
  cat("\n\n")
}
```

### Cyclins
#### Heatmap
```{r recat-dge-brightness-cyclins-heatmap, fig.width=7.8, fig.asp = 0.8}
cyclin_genes <- grep("^CCN[ABCDE][0-9]$", rowData(se)$gene_name)
# Extract matrix of expression
cc_mat <- assay(se, 'logcounts')[cyclin_genes, ]
rownames(cc_mat) <- rowData(se)[cyclin_genes, ]$gene_name
cc_zmat <- t(apply(cc_mat, 1, scale, center = TRUE, scale = TRUE))
colnames(cc_zmat) <- colnames(cc_mat)

# Generate Heatmap
htz <- Heatmap(
  cc_zmat,
  name= 'z-score',
  col = col_fun,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  
  top_annotation = column_ha,
  # cluster_columns = column_dend_colored,
  clustering_method_columns = 'ward.D2',
  column_split = use_annot$Cluster,
  cluster_column_slices = FALSE,
  column_title = NULL
)


draw(htz, 
     # heatmap_legend_side = "bottom", 
     annotation_legend_side = "bottom"
)
```

#### Distribution
```{r recat-dge-brightness-cyclins-exp}
cyclin_genes <- grep("^CCN[ABCDE][0-9]$", rowData(se)$gene_name)
use_data <- colData(se) %>% data.frame
i <- cyclin_genes[1]
for(i in cyclin_genes) {
  use_data$gene_exp <- assay(se, 'logcounts')[i, ]
  gene_name <- rowData(se)[i, ]$gene_name
  res <- use_data %>% 
    ggplot(aes(cluster, gene_exp, fill = cluster)) +
    geom_violin(width=0.9, scale = 'width', linewidth = rel(0.4)) +
    stat_summary(data = use_data %>% filter(dormancy_status!='mixed_dormant'),
                 fun.data="median_q1q3", 
                 mult=1, 
                 geom="crossbar", 
                 width=0.08, 
                 color = alpha('black', alpha = 1),
                 fill = 'white'
                 ) +
    # geom_jitter(shape=16, width=0.1, alpha = 0.8, show.legend = FALSE) +
    # geom_dotplot(shape=16, binaxis='y', stackdir='center', dotsize=0.9, alpha = 0.8) +
    geom_quasirandom(shape = 16, alpha = 0.5, width=0.25, size = rel(0.8)) +
    scale_fill_manual(values = cluster_palette) +
    labs(
      title = gene_name,
      x = 'Clusters',
      y = paste(gene_name, "expression"),
      fill = '',
      caption = paste(
        'Y-axis: normalized expression in log scale',
        'Points represent individual observations',
        'Box represent Q25, median and Q75',
        sep = '\n'
      )
    ) + 
    guides(fill = 'none') +
    # theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) +
    stat_compare_means(
      comparisons = my_comparisons,
      label.y.npc = 1,
      label.x.npc = 'left',
      size = 0.8*geom_text_size
      )
  print(res)
  cat("\n\n")

}

```





## Differential expression analysis
```{r knitr-template-rnaseq-deg-edger, eval = TRUE}
se$condition <- se$dormancy_brightness
group_palette <- get_scater_palette('tableau10medium')[1:nlevels(se$condition)] %>% 
  set_names(levels(se$condition))

# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-deg-edger.Rmd")
  )
```
`r knit(text = src)`

## Functional enrichment analysis
```{r knitr-template-rnaseq-gse-omnibus, eval = TRUE}
# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-gse-omnibus.Rmd")
  )
```
`r knit(text = src)`



