---
title: "Analysis of timeseries of DTC counts"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  data: ./data/external/experiments/timeseries_dtc_counts.xlsx
---
  
## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)
```

Load packages
```{r load-libraries}
library(tidyverse)
# library(showtext)
# library(foreach)
# library(DT)
library(knitr)
library(magrittr)
library(kableExtra)
# library(cowplot)
# library(colorblindr)
# library(ggbeeswarm)
# library(arsenal)
# library(patchwork)
library(openxlsx)
library(ggbeeswarm)
library(scales)

```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
source("./configuration/rmarkdown/color_palettes.R")
```


Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Prepare color palettes
```{r color-palettes}
vector_palette <- c(mVenus = '#109939', mCherry = '#e4017d')
```

## Analysis of Br16

Load and process data
```{r load-dw-br16}
data <- read.xlsx(params$data, sheet = 'Br16')
data %<>% 
  dplyr::select(mouse_id, mVenus_pos_norm, mVenus_neg_norm, time_months) %>% 
  dplyr::rename(n_dtc_positive = mVenus_pos_norm, n_dtc_negative = mVenus_neg_norm, subject = mouse_id, time = time_months) %>% 
  mutate(
    total_dtcs = n_dtc_positive + n_dtc_negative,
    prop_positive = ifelse(total_dtcs > 0, n_dtc_positive / total_dtcs, NA)
  )
```

### Data analysis including 0 counts


Keeping rows with 0 total DTCs can be informative if we want to understand the overall trend of mVenus+ DTCs over time, including the cases where there are no DTCs. However, it is important to handle these cases appropriately in the analysis and visualization, as they can lead to a large number of zero proportions that may not be meaningful for understanding the relationship between time and the Percentage of mVenus+ DTCs. In this case, we will keep the rows with 0 total DTCs but we will set the Percentage of mVenus+ DTCs to 0 for these cases, which allows us to include them in the analysis without distorting the visualization or the fitted model.

```{r glm-keep0-conf-br16}
use_data <- data %>% mutate(prop_positive = ifelse(is.na(prop_positive), 0, prop_positive))
```

```{r glm-keep0-bubble_plot-br16}
ggplot(use_data, aes(x = time, y = prop_positive)) +
  # geom_point(aes(size = total_dtcs), alpha = 0.6, color = vector_palette[['mVenus']]) + # Bubble size represents total cells
  geom_beeswarm(aes(size = total_dtcs), alpha = 0.6, color = vector_palette[['mVenus']]) + # Bubble size represents total cells
  geom_smooth(aes(weight = total_dtcs), method = "glm", 
              method.args = list(family = "binomial"), 
              color = "red", se = TRUE) +
  scale_x_continuous(breaks = use_data$time %>% unique) +
  labs(
    title = "Percentage of mVenus+ DTCs",
    x = "Time",
    y = "Percentage of mVenus+ DTCs",
    size = "Total DTCs",
    caption = "Proportion calculated over the\ntotal number of DTCs (mVenus+ and mVenus-)"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


```

Statistical analysis. Here we use a Generalized Linear Model (GLM) with a binomial error distribution and logit link function to model the relationship between time and the counts of positive and negative DTCs. This is much more robust statistically than running a linear regression on the calculated proportions, because the Binomial GLM inherently understands that a proportion of 50/100 carries much more statistical weight than a proportion of 1/2.
```{r glm-keep0-glm-br16}
model <- glm(cbind(n_dtc_positive, n_dtc_negative) ~ time, 
             data = use_data, 
             family = binomial(link = "logit"))

# summary(model)
broom::tidy(model) %>% 
  mutate(
    OR = exp(estimate),
    p.value = format.pval(p.value, digits = 3)
  ) %>% 
  dplyr::select(term, OR, estimate, std.error, statistic, p.value) %>% 
  kbl(caption = 'GLM results', align = "lrrrrr") %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```





## Analysis of 1833

Load and process data
```{r load-dw-1833}
data <- read.xlsx(params$data, sheet = '1833')
data %<>% 
  dplyr::select(mouse_id, mCherry_pos_norm, mCherry_neg_norm, time_months) %>% 
  dplyr::rename(n_dtc_positive = mCherry_pos_norm, n_dtc_negative = mCherry_neg_norm, subject = mouse_id, time = time_months) %>% 
  mutate(
    total_dtcs = n_dtc_positive + n_dtc_negative,
    prop_positive = ifelse(total_dtcs > 0, n_dtc_positive / total_dtcs, NA)
  )
```


### Data analysis including 0 counts


Keeping rows with 0 total DTCs can be informative if we want to understand the overall trend of mCherry+ DTCs over time, including the cases where there are no DTCs (Proportion = 0). However, it is important to handle these cases appropriately in the analysis and visualization, as they can lead to a large number of zero proportions that may not be meaningful for understanding the relationship between time and the Percentage of mCherry+ DTCs. In this case, we will keep the rows with 0 total DTCs but we will set the Percentage of mCherry+ DTCs to 0 for these cases, which allows us to include them in the analysis without distorting the visualization or the fitted model.

```{r glm-keep0-conf-1833}
use_data <- data %>% mutate(prop_positive = ifelse(is.na(prop_positive), 0, prop_positive))
```

```{r glm-keep0-bubble_plot-1833}
ggplot(use_data, aes(x = time, y = prop_positive)) +
  # geom_point(aes(size = total_dtcs), alpha = 0.6, color = vector_palette[['mCherry']]) + # Bubble size represents total cells
  geom_beeswarm(aes(size = total_dtcs), alpha = 0.6, color = vector_palette[['mCherry']]) + # Bubble size represents total cells
  geom_smooth(aes(weight = total_dtcs), method = "glm", 
              method.args = list(family = "binomial"), 
              color = "red", se = TRUE) +
  scale_x_continuous(breaks = use_data$time %>% unique) +
  labs(
    title = "Percentage of mCherry+ DTCs",
    x = "Time",
    y = "Percentage of mCherry+ DTCs",
    size = "Total DTCs",
    caption = "Proportion calculated over the\ntotal number of DTCs (mCherry+ and mCherry-)"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


```

Statistical analysis. Here we use a Generalized Linear Model (GLM) with a binomial error distribution and logit link function to model the relationship between time and the counts of positive and negative DTCs. This is much more robust statistically than running a linear regression on the calculated proportions, because the Binomial GLM inherently understands that a proportion of 50/100 carries much more statistical weight than a proportion of 1/2.
```{r glm-keep0-glm-1833}
model <- glm(cbind(n_dtc_positive, n_dtc_negative) ~ time, 
             data = use_data, 
             family = binomial(link = "logit"))

# summary(model)
broom::tidy(model) %>% 
  mutate(
    OR = exp(estimate),
    p.value = format.pval(p.value, digits = 3)
  ) %>% 
  dplyr::select(term, OR, estimate, std.error, statistic, p.value) %>% 
  kbl(caption = 'GLM results', align = "lrrrrr") %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

